Covering Your Basis: Model Agnostic Probes of the Binary Black Hole Population

### 2022-09-26 src/tex/intro.tex
fix words

Observations of gravitational waves (GWs) from compact binary coalescences (CBCs) is becoming a regular occurrence, 
recently producing catalogs of 90 such detection \cite{GWTC1,gwtc2,GWTC3}. With this growing catalog size, we are steadily approaching a more complete  
understanding of the population properties CBCs, just 7 years after the first observation \cite{o1o2_pop,o3a_pop,o3b_astro_dist}. 
In the coming fourth observing run of the LIGO-Virgo-KAGRA collaboration detectors \cite{aLIGO,aVIRGO, KAGRA}, with the increased sensitivities, 
there could be a detection as often as every day. This large influx of incoming data will drastically increase what we can learn by combining all observations
in order to studying the compact object population. The history of these dense objects is imprinted on the properties of the population 
of CBCs, allowing us to learn about their formation and evolution by the last few seconds of gravitation radiation emitted. The population distibutions 
of merging compact objects inferred from gravitational-wave observations has been shown to be parituclar useful in order to, measure cosmological parameters \citep{Farr_2019HUB}, 
constrain modified gravitational wave propagation, constrain a running Planck mass, look for evidence of ultralight bosons through superradiance, 
constrain stellar nuclear reaction rates \cite{Farmer_2020}, look for primordial black holes \cite{Ng_2021}, and constrain the neutron star equation of state \bruce{Cite all points}. 
Through a better understanding of the mass, spin, and redshift distributions of compact binaries that will come with the increased catalog size, 
one can probe a wide range of different physical phenoma with even greater fidelity.

@@

In this paper we highlight how the use of basis-splines can be a useful non-parametric modeling approach to the astrophysical distributions of compact 
binaries. We show that with a particular choice of basis, one can efficiently model the mass and spin distributions of compact binaries in GWTC-3 with
hierarchical Bayesian inference. We discuss our results in the context of current literature on compact object populations and illustrate how our results are similar
to those produced with the more simpler and astrophysically motivated parametric models. This data-drien approach will be capable of finding more subtle features, that 
may not be expected, should they appear in future catalogs. In Section \ref{sec:basis_splines} we go over the background of basis splines, in Section \ref{sec:results} 
we present results using our basis spline model on the mass and spin distributions of BBHs in GWTC-3, followed by discussing results in their astrophysical context and concluding 
in Section \ref{sec:conclusion}.

>>>>

Observations of gravitational waves (GWs) from compact binary coalescences (CBCs) is becoming a regular occurrence, 
recently producing catalogs of 90 such detection \cite{GWTC1,gwtc2,GWTC3}. With this growing catalog size, we are steadily approaching a more complete  
understanding of the population properties CBCs, just 7 years after the first observation \cite{o1o2_pop,o3a_pop,o3b_astro_dist}. 
In the coming fourth observing run of the LIGO-Virgo-KAGRA collaboration detectors \cite{aLIGO,aVIRGO, KAGRA}, with the increased sensitivities, 
there could be a detection as often as every day \cite{LVK_prospects}. This large influx of incoming data will drastically increase what we can learn by combining all observations
in order to studying the compact object population. The history of these dense objects is imprinted on the properties of the population 
of CBCs, allowing us to learn about their formation and evolution by the last few seconds of gravitation radiation emitted. The population distibutions 
of merging compact objects inferred from gravitational-wave observations has been shown to be parituclar useful in order to, measure cosmological parameters \citep{Farr_2019HUB}, 
constrain modified gravitational wave propagation, constrain a running Planck mass, look for evidence of ultralight bosons through superradiance, 
constrain stellar nuclear reaction rates \cite{Farmer_2020}, look for primordial black holes \cite{Ng_2021}, and constrain the neutron star equation of state \bruce{Cite all points}. 
Through a better understanding of the mass, spin, and redshift distributions of compact binaries that will come with the increased catalog size, 
one can probe a wide range of different physical phenoma with even greater fidelity.

@@

Polynomial splines have been applied to success across different areas of gravitational-wave astronomy. They have been used to model the gravitational-wave data noise spectrum, 
detector calibration uncertainties, coherent gravitational waveform deviations, and modulations to a power law mass distribution \cite{Littenberg_2015,Edwards_2018,B_Farr_etal_2014,Edelman_2021,Edelman_2022ApJ}
In this paper we highlight how the use of basis-splines can provide a powerful non-parametric modeling approach to the astrophysical distributions of compact 
binaries. We illustrate how one can efficiently model the mass and spin distributions of compact binaries in GWTC-3 with basis splines to infer population properties using 
hierarchical Bayesian inference. We discuss our results in the context of current literature on compact object populations and how this method complements the simpler lower dimensional parametric models. 
This data-drien approach will be capable of finding more subtle features, that may not be expected, should they appear in future catalogs. 
In Section \ref{sec:basis_splines} we go over the background of basis splines, in Section \ref{sec:results} we present results using our basis 
spline model on the mass and spin distributions of BBHs in GWTC-3, followed by discussing results in their astrophysical context and concluding in Section \ref{sec:conclusion}.
